EuPPollNet: A European database of plant-pollinator networks

### 2024-05-24 Manuscript/EuPPollNet_Ms.qmd
paragraph 4 more or less on the way now

Here, we present the EuPPollNet database, an acronym derived from European plant-pollinator networks, which contains harmonized information on interaction data of plants and pollinators at European level. The primary focus of the pollinator taxonomic groups is on the main orders of entomofauna that visit and pollinate flowering plants in Europe. These include insect species from the orders Hymenoptera, Diptera, Lepidoptera, and Coleoptera [@potts2015], which comprise almost the totality of recorded interactions in the EuPPollNet database (`r format(round(main_orders_percent, 2), nsmall = 2)`%). To understand the scope of the database, we examine the taxonomic and sampling coverage of the different plant and pollinator species at European level with the help of the most up-to-date species checklists and rarefaction analyses. In addition, we investigate how two key structural metrics of plant-pollinator networks, such as connectance and nestedness, change across the latitudinal range of studies and bioclimatic regions. Finally, we also examine whether plant-pollinator networks fulfill the expectation of being more nested than null expectations. EuPPollNet aims to cover a wide range of taxonomic groups and habitats, while also providing other variables of interest that allow better control for the ecological context and sampling methods. In addition, EuPPollNet offers a transparent and accessible workflow of its data management and species harmonization that allows it to be reused and keep building on it over time. This database provides a large number of community-level networks with curated and harmonized data, distinguishing it from other currently available resources that contain plant-pollinator interactions. We expect that this database will help to evaluate macro-ecological processes and current gaps of plant-pollinator interactions at European level.

@@

Bees conducted the majority of the interactions at the metaweb level, but their relative relevance changed across habitats and bioclimatic regions. For instance, plant-pollinator communities in the Mediterranean were dominated by bees, while communities in Alpine or Boreal regions were especially fly species rich or fly-dominated. These patterns are consistent with our current understanding of bee diversity, which peaks in dry or temperate areas [@orr2021; @leclercq2023]; and with the fact that colder environments (i.e., altitude and latitude wise) harbor a larger fraction of fly pollinators compared to other taxa [@elberling1999; @lefebvre2018]. In addition, beetles were only species rich in the Mediterranean region. Although the networks from the database are visitation networks and do not capture pollinator efficiency [@ballantyne2015], the high proportion of beetles as floral visitors in the Mediterranean provides further support for their role as pollinators in this region [@herrera2019; @leon2022]. The number of butterfly species and interactions were relatively low compared to the other taxa. While Europe contains fewer butterfly species than other regions of the world [@ollerton2017], their relevance as pollinators is likely underestimated within this database. This is because a large fraction (~40%) of studies did not sample butterflies, and conventional sampling methods for monitoring other insect pollinators (e.g., bees or flies) may be inadequate for sampling plant-butterfly interactions [@isaac2011]. Honey bees were present in `r round(apis_percent_networks, 1)`% of networks and conducted in average a third of the total interactions per network. This proportion is higher that the one found when considering only natural communities at a global scale [~13%\; @hung2018], which highlights the need of further work evaluating how honey bees and their management impact natural pollinator communities.

Although Europe contains a much larger number of flowering plants than pollinator species (~4 to 1 ratio), the observed number of pollinator species almost doubled the one of plants in the database. This could be explained by the fact that all networks are phytocentric, resulting in sampling bias towards pollinator species [@jordano2016; @vizentin2018]. While animal-centred sampling is likely to increase the plant:pollinator species ratio [e.g., @encinas2023], the spatial scale and environmental context of the sampled communities will also influence their observed diversity, especially given the ability to move of pollinators and the sessile nature of plants. As expected [@jordano1987], we found that network connectance decreased exponentially with the number of species. Overall, latitude did not explain major changes in residual connectance or normalised nestedness of the plant-pollinator networks. Consistent with @olesen2002, residual connectance was lower at higher latitudes, while normalised nestedness, which is negatively correlated with residual connectance, increased towards higher latitudes. Given their correlation, the potential ecological explanations that can help understanding these structural patterns can be similar. For instance, networks at lower latitudes in Europe are exposed to higher temperatures and are bee-dominated, which can result in higher visitation rates [@arroyo1985; @classen2015; @herrera2019]; and, the level of pollinator generalization seems to be higher at lower latitudes [@schleuning2012]. Therefore, these different factors can increase the possible number of connections that can be established between plants and pollinators for a given network size, resulting in more connected networks but less nested ones in comparison to networks at higher latitudes. In addition, while one third of networks were more nested than expected by chance, two thirds of them did not show statistical difference with null expectations. This goes against the common view that all plant-pollinator networks are nested [@bascompte2003], and supports the idea that despite a strong tendency towards it, being nested is not the rule in this type of mutualism.

Despite this database covers a wide range of habitats across Europe, still contains major geographical gaps such as the East of Europe or the Mediterranean region.

Maybe some refs about the above phrase?

Add that

>>>>

Here, we present the EuPPollNet database, an acronym derived from European plant-pollinator networks, which contains harmonized information on interaction data of plants and pollinators at European level. The primary focus of the pollinator taxonomic groups is on the main orders of entomofauna that visit and pollinate flowering plants in Europe. These include insect species from the orders Hymenoptera, Diptera, Lepidoptera, and Coleoptera [@willemstein1987; @potts2015], which comprise almost the totality of recorded interactions in the EuPPollNet database (`r format(round(main_orders_percent, 2), nsmall = 2)`%). To understand the scope of the database, we examine the taxonomic and sampling coverage of the different plant and pollinator species at European level with the help of the most up-to-date species checklists and rarefaction analyses. In addition, we investigate how two key structural metrics of plant-pollinator networks, such as connectance and nestedness, change across the latitudinal range of studies and bioclimatic regions. Finally, we also examine whether plant-pollinator networks fulfill the expectation of being more nested than null expectations. EuPPollNet aims to cover a wide range of taxonomic groups and habitats, while also providing other variables of interest that allow better control for the ecological context and sampling methods. In addition, EuPPollNet offers a transparent and accessible workflow of its data management and species harmonization that allows it to be reused and keep building on it over time. This database provides a large number of community-level networks with curated and harmonized data, distinguishing it from other currently available resources that contain plant-pollinator interactions. We expect that this database will help to evaluate macro-ecological processes and current gaps of plant-pollinator interactions at European level.

@@

Bees conducted the majority of the interactions at the metaweb level, but their relative relevance changed across habitats and bioclimatic regions. For instance, plant-pollinator communities in the Mediterranean were dominated by bees, while communities in Alpine or Boreal regions were especially fly species rich or fly-dominated. These patterns are consistent with our current understanding of bee diversity, which peaks in dry or temperate areas [@orr2021; @leclercq2023]; and with the fact that colder environments (i.e., altitude and latitude wise) harbor a larger fraction of fly pollinators compared to other taxa [@elberling1999; @lefebvre2018]. In addition, beetles were only species rich in the Mediterranean region. Although the networks from the database are visitation networks and do not capture pollinator efficiency [@ballantyne2015], the high proportion of beetles as floral visitors in the Mediterranean provides further support for their role as pollinators in this region [@herrera2019; @leon2022]. The number of butterfly species and interactions were relatively low compared to the other taxa. While Europe contains fewer butterfly species than other regions of the world [@ollerton2017], their relevance as pollinators is likely underestimated within this database. This is because a large fraction of studies (~40%) did not sample butterflies, and conventional sampling methods for monitoring other insect pollinators (e.g., bees or flies) may be inadequate for sampling plant-butterfly interactions [@isaac2011]. Honey bees were present in `r round(apis_percent_networks, 1)`% of networks and conducted in average a third of the total interactions per network. This proportion is higher that the one found when considering only natural communities at a global scale [~13%\; @hung2018], which highlights the need of accounting for the potential negative impact of honey bees to wild pollinator communities when managing pollinator communities in Europe [@steffan2000; @herrera2020].

Although Europe contains a much larger number of flowering plants than pollinator species (~4 to 1 ratio), the observed number of pollinator species almost doubled the one of plants in the database. This could be explained by the fact that all networks are phytocentric, resulting in sampling bias towards pollinator species [@jordano2016; @vizentin2018]. While animal-centred sampling is likely to increase the plant:pollinator species ratio [e.g., @encinas2023], the spatial scale and environmental context of the sampled communities will also influence their observed diversity, especially given the ability to move of pollinators and the sessile nature of plants. As expected [@jordano1987], we found that network connectance decreased exponentially with the number of species. Overall, latitude did not explain major changes in residual connectance or normalised nestedness of the plant-pollinator networks. Consistent with @olesen2002, residual connectance was lower at higher latitudes, while normalised nestedness, which is negatively correlated with residual connectance, increased towards higher latitudes. Given their correlation, the potential ecological explanations that can help understanding these structural patterns can be similar. For instance, networks at lower latitudes in Europe are exposed to higher temperatures and are bee-dominated, which can result in higher visitation rates [@arroyo1985; @classen2015; @herrera2019]; and, the level of pollinator generalization seems to be higher at lower latitudes [@schleuning2012]. Therefore, these different factors can increase the possible number of connections that can be established between plants and pollinators for a given network size, resulting in more connected networks but less nested ones in comparison to networks at higher latitudes. In addition, while one third of networks were more nested than expected by chance, two thirds of them did not show statistical difference with null expectations. This goes against the common view that most plant-pollinator networks are nested [@bascompte2003], and supports the idea that despite a strong tendency towards it, being nested is not the rule fo this large set of plant-pollinator networks.

Despite this database covering a wide range of habitats across `r country_number` European countries, some areas such as Eastern Europe and the Mediterranean region are evidently under-represented. This reflects the lack of plant-pollinator community data in highly relevant regions of pollinator diversity [REFs] and could help explain the limited taxonomic coverage found of plant and pollinator species at the European level. This is especially relevant as Eastern Europe is expected to continue developing and use their vast natural resources [@bennett2018] and the Mediterranean region is likely to be severely impacted by climate change [@jaworski2022; @duchenne2020; @pareja2023]. Although some of the most well studied countries in Europe (e.g., Belgium or The Netherlands) seem to have suffered most of the plant and pollinator biodiversity loss at the end of the 20th century [@carvalheiro2013], plant-polllinator communities in Europe and across the globe still suffer from major threads such as climate change [REFs], land use change [REFs], or the introduction of alien species [@vila2009; @vanbergen2018].

Improve this idea but something like this: -make it more discussion style-

Recording their interactions and understanding plant-pollinator community assembleges now is essential to understand the trends of biodiversity, as shown by the importance of historical or time series datasets.
